Tools for Identifying Partisan Gerrymandering (PA)

### 2019-06-13 Latex/_maintext.tex
Update _maintext.tex
revising text and code
Updated LaTeX files

% -- SECTION -- SECTION -- SECTION -- SECTION -- SECTION -- SECTI %

@@

        \footnote{The trial court considered a variety of factors offered in expert witness testimony, including evidence about the map‘s political impact. The trial court’s findings of fact reveal that the 2011 map exhibited clearly a multiplicity of indicia of partisan gerrymandering. In his review, the trial judge found, in essence, that the legislative process through which the plan had been chosen demonstrated the indicia of a partisan gerrymander, that the weird shapes of a number of the districts demonstrated the indicia of a partisan gerrymander, that the dismemberment of municipalities, townships and counties in the state demonstrated the indicia of a partisan gerrymander, that the cracking and packing of Democrat-leaning voters demonstrated the indicia of a partisan gerrymander, and that the frozen 13R-5D results over the course of three elections in a state that is highly politically competitive statewide demonstrated the indicia of a partisan gerrymander. And the trial judge also found that none of these features of the map could be explained as necessitated by the electoral geography and demography of the state, though he suggested the possibility that consideration such as incumbency protection might have mattered. However, the trial court then, rather regretfully, concluded that it did not have a legal basis to declare the plan unconstitutional under either the Pennsylvania State Constitution or under then existing federal law precedents. Instead, it expressed a pious hope that the U.S. Supreme Court, in its review of the appeals of lower court cases, such as the 2017 case of \href{https://www.brennancenter.org/sites/default/files/legal-work/16-1161_Opinion.pdf}{\textit{Gill v. Whitford 585 U. S. \_\_\_ (2018)}.}, which had overturned as a partisan gerrymander a Wisconsin legislative map, would finally adopt a (manageable) standard for identifying unconstitutional partisan gerrymandering that all courts could apply. The Pennsylvania Supreme Court, for all practical purposes, adopted in \textit{toto} the factual finding of the trial court magistrate, Judge Brobson, but came to a diametrically opposite legal conclusion about the constitutionality of the congressional plan and the legal grounding for that finding.}

@@

    Having invalidated the \textit{2011 Enacted} map, the Pennsylvania Supreme Court offered deferrence to the State to draw a remedy plan. The Pennsylvania legislature failed to enact a lawful remedy plan of its own due mostly to divided government that led to disagreement as to how to proceed between the newly elected Democratic governor and the still Republican-controlled legislature. Had such a plan been passed (and signed), the Court would have reviewed its constitutionality. In the absence of a legislatively enacted remedial plan, the Court invited submissions of plans from the parties to the lawsuit, as well the public at large, to submit remedy plans; and it provided guidance as to the features of the plan in which it was most interested. The Court Order in February 2018 inviting submissions of feasible maps mentioned only good government criteria: minimizing the number of county, township and municipality splits, and maximizing district compactness under several different measures of compactness. It made no mention of incumbent protection, nor voting rights act protections, nor of protecting communities of interests, nor did it mention political outcome projections.

@@

        \footnote{Pennsylvania Supreme Court Judges are elected or retained to ten-year terms in elections with party labels on the ballot. Judges with Democratic affiliations now constitute the majority of the Court In 2011 this was not the case.}

@@

    In contrast to the \textit{LWV} finding that the 2011 Pennsylvania congressional map was unconstitutional under state law, a 2017 challenge to the Pennsylvania congressional map under the Elections Clause of the U.S. Constitution was rejected by a three-judge federal court by a vote of 2-1. Because the \textit{LWV} decision mooted the federal litigation, the federal challenge was not appealed to the U.S. Supreme Court. However, a 2018 remand of a Wisconsin partisan gerrymandering case, \textit{Gill v. Whitford} has, to a limited extent, provided clues about the current [ca. May 2019] state of federal partisan gerrymandering jurisprudence than anything before it.

@@

        \footnote{The \textit{Gill} remand decision still leaves open the question that has haunted opponents of partisan gerrymandering for the last thirty plus years, namely `When, if ever, will the U.S. Supreme Court actually declare a redistricting map unconstitutional?'.  The potential for such a finding of unconstitutionality is laid out in \textit{Davis v. Bandemer, 478 U.S. 109 (1986)}, which declared partisan gerrymandering justiciable, but under which the Supreme Court has never yet (as of May 2019) struck down a plan as an unconstitutional partisan gerrymander. Despite the 7-2 decision on the Wisconsin remand, which rejected the view that the Court should simply bow out of the process of reviewing maps as possible unconstitutional partisan gerrymanders, the retirement of Justice Kennedy and his replacement by a more conservative Justice creates even greater uncertainty about whether, in the lifetime of the present authors, there will ever be a U.S. Supreme Court majority that rejects some challenged district as an unconstitutional partisan gerrymander. We will, relatively soon, have the answer to that question due to the inevitable appeals of partisan gerrymandering cases from North Carolina and Maryland that have already had oral argument on March 26, 2019 before the U.S. Supreme Court (\textit{Rucho, et. al}. v. \textit{Common Cause, et. al} [1:16-CV-1026]) and \textit{Lamone, et. al.} v. \textit{Benisek, et. al.} [1:13-cv-03233-JKB]), with decisions expected in June 2019. In addition, there is a partisan gerrymandering challenge in Michigan, \textit{League of Women Voters of Michigan} v. \textit{Johnson} that will certainly be appealed to the U S. Supreme Court, as will the next lower court decision in \textit{Gill} itself.}

@@

    Districts that have been \textit{packed} are those which have had voters from one partisan leaning grouped together so as to waste votes of the group because they win the district by an overwhelming majority, thus preventing those  excess votes from being used more efficiently in other districts. Districts that have been \textit{cracked} are ones where voters of one partisan leaning have had their share of the district’s voting strength reduced so as to make it impossible (or unlikely) that members of this group will be able to carry the district. The standard for unconstitutional gerrymandering at the federal level -- if there is ever is one -- can thus be expected to be very different from that offered by the Pennsylvania Supreme Court, since \textit{LWV} relies on data about the features and effects of the plan as a whole, while the US Supreme Court appears to be insisting on a district-specific test of packing and cracking. 

@@

    Although only directly applicable in Pennsylvania, \textit{LWV} is, we believe, nonetheless the most important partisan gerrymandering case yet decided in a final and definitive fashion this decade. 

@@

    Because the US Constitution is the minimum rights provided to all citizens, states offer additional protections. Language virtually identical to the provision of the state constitution that the Pennsylvania Supreme Court relied on is found in the constitutions of another dozen states, and another thirteen have very similar language \citep{Douglas2014_RightToVote}.

@@

    However, as we show later, some states in which possible state law challenges could be made will be insulated from those challenges by factors such as the nature of institutional control over the redistricting process in the state, or political gridlock in the state courts. Nonetheless, if a state court does follow the Pennsylvania Supreme Court lead, we view the expected effects on limiting partisan gerrymandering to be potentially greater than we might expect under federal law, since they would involve the overturning of whole plans, not just individual districts -- although, in practice, a finding of overall unconstitutionality is likely to lead to changes only in the most egregiously gerrymandered districts and some of the districts adjacent to them. Unlike all federal courts that have ruled on partisan gerrymandering to date, the Pennsylvania opinion does not require proof of intent to discriminate, which could make proving a violation easier under state law than under any likely federal standard. Moreover, although the Pennsylvania Supreme Court opinion does not require evidence of partisan effects, and emphasizes failure to comply with good government criteria, we believe that the \textit{LWV} opinion can be ``tweaked" to address what \citet{GrofmanCervas2018_WashPo} in their on-line essay in the \textit{Washington Post} refer to as \textit{stealth gerrymanders}, i.e., plans that satisfy good government redistricting criteria to a reasonable degree but are nonetheless carefully crafted invidious partisan gerrymanders in terms of their expected political impact. The reason for our optimism in this respect are the findings in the opinion that do address partisan effects of the Pennsylvania congressional map. 

@@

    In addition to the rejected 2011 and the court-ordered 2018 map, we also address the partisan and other features of two maps offered to the Pennsylvania Court as a proposed remedy for constitutional violation found in the 2011 map: the \textit{joint legislative} plan offered by Republican legislators from both chambers, and the congressional map proposed by the Democratic governor (\textit{Gov. Wolf}). Here we focus on the question of whether either of these plans was a \textit{stealth gerrymander} done to aid the party drawing the map.

@@

% -- SUBSECTION -- SUBSECTION -- SUBSECTION -- SUBSECTION -- SUBS %
    Avoiding county splits is incorporated into the language of many state constitutions \citep{Grofman1985} and can be seen as a way of guarding against gerrymandering by making it more difficult to draw partisan maps. There are a total of 67 counties in PA, and the state is allocated 18 congressional districts.  Some county splits are necessitated by the size of the county population relative to the ideal population of a congressional district (705,687).  Allegheny and Montgomery must each be split at least twice, while Philadelphia must be split at least three times. If no set of counties exactly equals the ideal congressional population, every districting plan requires a minimum of $n-1$ splits (where $n$ equals the number of congressional districts). The Pennsylvania Supreme Court found that an unusually and egregiously large number of counties were split in the 2011 congressional map, and they found this fact deeply troubling.

@@

@@

       \footnote{There is a debate about how to count split counties, which was evident when different parties in the case counted different total county splits for the same plan. One way to count splits to count the total number of counties that are in at least two districts. The other way is to count the number of times a county is split, i.e., the number of separate pieces into which the county is divided. In the first counting method, a county can have portions of itself found in 18 different districts, but have this counted as only one county spit. Under the second method, this wholesale splitting of the county would be counted as 18 splits. While both methods provide useful information, we regard the second as more informative and it is the one reported in Table 1. However, we must also be alert for divisions of a county into two or more discontiguous pieces found within the same district. We call this “fracking,” to provide a term similar sounding to ‘packing’ and ‘cracking.’ We find “fracking” a particularly egregious problem since there is no need for counties to be fracked, and fracking may be used as a gerrymandering tool.  Instances where counties  have multiple discontiguous pieces located in a single district are not discussed in the Pennsylvania Supreme Court ruling on the unconstitutionality of the 2011 Pennsylvania congressional plan. However, our analysis of the 2011 plan reveals several cases of fracking: two of the most egregious are districts 14 and 18 in Allegheny County (Pittsburgh). This concept has also been addressed by the US District Court in North Carolina (\textit{Common Cause v. Rucho} No 1:16-CV-1026 (U.S. District Court, Middle District of North Carolina, 2018, slip op at p. 105 [p. 194])). }

@@

@@

    \footnote{We find counties that are split into multiple pieces particularly egregious. ``\textit{Fracking}" occurs when one district is split into two or more non-contiguous pieces of a single county. It was not mentioned in the \textit{LWV} decision, but too can be used as an indicia of partisan (and racial) manipulation. We call this ``\textit{fracking}" to parallel the terminology of other districting terms of ``\textit{cracking}" and ``\textit{packing}". Examples of ``fracking" in the 2011 enacted map include districts 14 \& 18 in Allegheny County (Pittsburgh).}

@@

@@

    To give a more intuitive feel for how egregiously ill-compact the 2011 map is compared to the other alternatives we are considering, we show all four maps in the same scale, in a gray and white format. What is visually apparent from these maps is how aesthetically ``ugly" the 2011 map is compared to all of its alternatives. But it also clear that three remedy maps that are similar in having relatively few county cuts nonetheless can look quite different in terms of how each configures districts.

@@

% -- FIGURE -- FIGURE -- FIGURE -- FIGURE -- FIGURE -- FIGURE --   % 
% Maps Of The Proposed Pennsylvania Congressional Districts
        \
% -- END FIGURE -- END FIGURE -- END FIGURE -- END FIGURE -- END FI %

@@

@@

@@

    There are other indicators of statewide partisan gerrymandering effects that also provide a compelling picture of the 2011 map as an (enduring) partisan gerrymander. There are several metrics that have been proposed to look at the degree to which there is asymmetry in seats-votes relationships that might be indicative of partisan gerrymandering: (1) \textit{partisan bias} \citep{Tufte1973}, (2) the \textit{mean-minus-median} gap \citep{Mcdonald_Best_2015_ELJ} (3) the \textit{efficiency gap} \citep{Stephanopoulos2014_UofChicagoLaw}, and most recently (4) declination \citep{Warrington2018}. Each of these measures can be directly calculated for the 2011 map based on district election outcomes. 

@@

    \footnote{Although the basic idea goes back at least as far as \citet{Dahl1956}, the modern \textit{locus classicus} in political science for the idea of partisan bias(and the parallel concept of responsiveness/swing ratio) is \citet{Tufte1973}, with important recent work done by Gary King (see e.g., \citep{GelmanKing1994_unifiedAJPS}; and review in \citet{GrofmanKing2007}). A closely related line of research is found in the political geography literature (with \textit{locus classicus} \citet{Brookes1959, Brookes1960} -- with some further methodological improvements by scholars such as \citet{Johnston1994, Rossiter1997} and \citet{Johnston2002}.}   

@@

    \footnote{Where to center the \textit{seats/votes} curve is an unresolved issue in the literature. In \citet{GelmanKing1994_unifiedAJPS} \texttt{JudgeIt} program, the choice is to center the election at the average Democratic vote share over recent elections. \citet{Kastellec_et_al_2008_PS}, similarly, simulates data for average vote shares between 45\% and 55\%. We checked for consistency between the methods by running the numbers in three ways: centering the simulated outcomes at an average election vote share, centering the simulated outcomes at a tied 50/50 election, and centering the simulated outcomes at the actual election results in the particular election. All three variations provide similar estimates, as do the results from \texttt{JudgeIt}. Readers should note that with only 18 districts, too much weight should not be place exact bias coefficients reported [see the discussion in \citet{Browning_King_1987_seats_votes}]; but the relative magnitude of the bias values across the different plans is fully congruent with the evaluations of probable partisan gerrymandering yielded by the other measures. Because of the limited number of districts in the plans, the same caution is urged for all the measures of gerrymandering reported in this essay.} 

@@

    The \textit{Efficiency Gap} is calculated as defined in \citet{Stephanopoulos2014_UofChicagoLaw}, where all the party’s votes are wasted if they lose the district, and all the winner’s votes over 50\% are wasted. The difference between each party’s wasted votes is then divided by the total votes cast to produce the \textit{Efficiency Gap}, with a value of zero denoting what is regarded as ideal. As noted in \citet[][p. 13]{Best2018}, this is equivalent to taking an aggregate responsiveness (swing ratio) of two as ideal. 

@@

    Among the more recent measures of gerrymandering is the declination. Declination is an astronomical term referring to the angle on a compass, with a Latin root meaning ``a bending away". For gerrymandering, it uses the angles created from ordering all of the districts by vote share, computing the mean vote-share for each party separately for the seats they won, and then comparing the differences in their distribution. To compare, a line is draw from the mean vote-share to the 0.5 line separately for each party. The angle formed from the different slopes of the line is called the declination. An angle indicates whether a plan is a gerrymandering, under the logic that vote-shares should be distributed in some uniform manner across districts. If partisan are \textit{packed}, the angle will be greater. All values range between -1 and 1. Though we have defined the vote-share in terms of Republican outcomes, we adjusted the values such that \underline{negative} values are favorable to the Democrats. Further information can be found in \citet{Warrington2018}.  

@@

    Rebuttal to any claims that the partisan asymmetries observed in the consistent outcomes of the 2011 map can be attributed primarily to differential concentration of party voting strengths, or to chance, have been decisively refuted by the expert witness testimony in the case (see esp. the simulations of potential good government maps done by Professor Chen, discussed in detail in the Pennsylvania Supreme Court opinion in the case, and on which the Court placed great reliance). With respect to the durability of the effects of the 2011 map, we have already pointed out that this plan yielded unchanged 13R-5D results for three elections in a row. To help explain this over time consistency we note that the 2011 Map has primarily safe seats (13 of 18 won with greater than a 10\% margin). But, indicative of the pro-Republican gerrymander achieved with the 2011 map, those safe seats are distributed in a highly asymmetric fashion, with nine essentially safe seats for Republicans and only four for Democrats.

@@

    \begin{center}    \textbf{US House Election Summaries (PA 2004-2016)} \par \end{center}
    \begin{center}   \textbf{[ INSERT TABLE \ref{tab:congsum} ABOUT HERE ]} \end{center}

@@

		\footnote{We can only validate our measure for the 2016 election and only for the 2011 enacted plan. For all the remedial plans we are analyzing, the composite is our ``best guess". We have to make some assumptions about state and district inter-election swings. For simplicity, we assume that all districts have a national shift equal to the state-level swing, ie, difference from the 2016 elections. Later in the paper, we conduct simulated elections that vary the district level variability.}

@@

		\footnote{Whether we take the sum of the five elections or average the five elections, both methods deliver the same relative distributions.}

@@

    Additionally, each district's outcome is identical to the 2016 congressional district results. 

@@

		\footnote{If we center the simulations at the 2012-2016 average statewide Republican vote of 53.3\% of the vote, we again get remarkably consistent estimates [0.09, 0.19, 0.05, 0.46 ].}

@@

    Such clauses are found in a number of the states widely regarded as having the worst congressional gerrymandering. 

>>>>

% █▀▀ █▀▀ █▀▀ ▀▀█▀▀ ░▀░ █▀▀█ █▀▀▄
% ▀▀█ █▀▀ █░░ ░░█░░ ▀█▀ █░░█ █░░█
% ▀▀▀ ▀▀▀ ▀▀▀ ░░▀░░ ▀▀▀ ▀▀▀▀ ▀░░▀

@@

        \footnote{The trial court considered a variety of factors offered in expert witness testimony, including evidence about the map‘s political impact. The trial court’s findings of fact reveal that the 2011 map exhibited clearly a multiplicity of indicia of partisan gerrymandering. In his review, the trial judge found, in essence, that the legislative process through which the plan had been chosen demonstrated the indicia of a partisan gerrymander, that the weird shapes of a number of the districts demonstrated the indicia of a partisan gerrymander, that the dismemberment of municipalities, townships and counties in the state demonstrated the indicia of a partisan gerrymander, that the cracking and packing of Democrat-leaning voters demonstrated the indicia of a partisan gerrymander, and that the frozen 13R-5D results over the course of three elections in a state that is highly politically competitive statewide demonstrated the indicia of a partisan gerrymander. The trial judge also found that none of these features of the map could be explained as necessitated by the electoral geography and demography of the state, though he suggested the possibility that consideration such as incumbency protection might have mattered. However, the trial court then, rather regretfully, concluded that it did not have a legal basis to declare the plan unconstitutional under either the Pennsylvania State Constitution or under then existing federal law precedents. Instead, it expressed a pious hope that the U.S. Supreme Court, in its review of the appeals of lower court cases, such as the 2017 case of \href{https://www.brennancenter.org/sites/default/files/legal-work/16-1161_Opinion.pdf}{\textit{Gill v. Whitford 585 U. S. \_\_\_ (2018)}.}, which had overturned as a partisan gerrymander a Wisconsin legislative map, would finally adopt a (manageable) standard for identifying unconstitutional partisan gerrymandering that all courts could apply. The Pennsylvania Supreme Court, for all practical purposes, adopted in \textit{toto} the factual finding of the trial court magistrate, Judge Brobson, but came to a diametrically opposite legal conclusion about the constitutionality of the congressional plan and the legal grounding for that finding.}

@@

    Having invalidated the \textit{2011 Enacted} map, the Pennsylvania Supreme Court offered deference to the State to draw a remedy plan. The Pennsylvania legislature failed to enact a lawful remedy plan of its own due mostly to divided government that led to disagreement as to how to proceed between the newly elected Democratic governor and the still Republican-controlled legislature. Had such a plan been passed (and signed), the Court would have reviewed its constitutionality. In the absence of a legislatively enacted remedial plan, the Court invited submissions of plans from the parties to the lawsuit, as well the public at large, to submit remedy plans; and it provided guidance as to the features of the plan in which it was most interested. The Court Order in February 2018 inviting submissions of feasible maps mentioned only good government criteria: minimizing the number of county, township and municipality splits, and maximizing district compactness under several different measures of compactness. It made no mention of incumbent protection, nor voting rights act protections, nor of protecting communities of interests, nor did it mention political outcome projections.

@@

        \footnote{Pennsylvania Supreme Court Judges are elected or retained to ten-year terms in elections with party labels on the ballot. Judges with Democratic affiliations now constitute the majority of the Court. In 2011 this was not the case.}

@@

    In contrast to the \textit{LWV} finding that the 2011 Pennsylvania congressional map was unconstitutional under state law, a 2017 challenge to the Pennsylvania congressional map under the Elections Clause of the U.S. Constitution was rejected by a three-judge federal court by a vote of 2-1. Because the \textit{LWV} decision mooted the federal litigation, the federal challenge was not appealed to the U.S. Supreme Court. However, a 2018 remand of a Wisconsin partisan gerrymandering case, \textit{Gill v. Whitford} (referred to interchangeably as \textit{Gill}) has, to a limited extent, provided clues about the current [ca. June 2019] state of federal partisan gerrymandering jurisprudence than anything before it.

@@

        \footnote{The \textit{Gill} remand decision still leaves open the question that has haunted opponents of partisan gerrymandering for the last thirty plus years, namely `When, if ever, will the U.S. Supreme Court actually declare a redistricting map unconstitutional?'.  The potential for such a finding of unconstitutionality is laid out in \textit{Davis v. Bandemer, 478 U.S. 109 (1986)}, which declared partisan gerrymandering justiciable, but under which the Supreme Court has never yet (as of June 2019) struck down a plan as an unconstitutional partisan gerrymander. Despite the 7-2 decision on the Wisconsin remand, which rejected the view that the Court should simply bow out of the process of reviewing maps as possible unconstitutional partisan gerrymanders, the retirement of Justice Kennedy and his replacement by a more conservative justice creates even greater uncertainty about whether, in the lifetime of the present authors, there will ever be a U.S. Supreme Court majority that rejects some challenged district (or plan) as an unconstitutional partisan gerrymander. We will relatively soon have the answer to that question due to the inevitable appeals of partisan gerrymandering cases from North Carolina and Maryland that have already had oral argument on March 26, 2019 before the U.S. Supreme Court (\textit{Rucho, et. al}. v. \textit{Common Cause, et. al} [1:16-CV-1026]) and \textit{Lamone, et. al.} v. \textit{Benisek, et. al.} [1:13-cv-03233-JKB]), with decisions expected in June 2019. In addition, there is a partisan gerrymandering challenge in Michigan, \textit{League of Women Voters of Michigan} v. \textit{Johnson} that will certainly be appealed to the U.S. Supreme Court, as will the next lower court decision in \textit{Gill} itself.}

@@

    Districts that have been \textit{packed} are those which have had voters from one partisan leaning grouped together so as to waste votes of the group because they win the district by an overwhelming majority, thus preventing those excess votes from being used more efficiently in other districts. Districts that have been \textit{cracked} are ones where voters of one partisan leaning have had their share of the district’s voting strength reduced so as to make it impossible (or unlikely) that members of this group will be able to carry the district. The standard for unconstitutional gerrymandering at the federal level -- if there is ever is one -- can thus be expected to be very different from that offered by the Pennsylvania Supreme Court, since \textit{LWV} relies on data about the features and effects of the plan as a whole, while the U.S. Supreme Court appears to be insisting on a district-specific test of packing and cracking. 

@@

    Although only directly applicable in Pennsylvania, \textit{LWV} is, we believe nonetheless, the most important partisan gerrymandering case yet decided in a final and definitive fashion this decade. 

@@

    Because the U.S. Constitution provides for the minimum rights provided to all citizens, state laws and constitutions offer additional protections. Language virtually identical to the provision of the PA state constitution that the Pennsylvania Supreme Court relied on is found in the constitutions of another dozen states, and another thirteen have very similar language \citep{Douglas2014_RightToVote}.

@@

    Some states in which possible state law challenges could be made will be insulated from those challenges by factors such as the nature of institutional control over the redistricting process in the state, or political gridlock in the state courts. Nonetheless, if a state court does follow the Pennsylvania Supreme Court lead, we view the expected effects on limiting partisan gerrymandering to be potentially greater than we might expect under federal law, since they would involve the overturning of whole plans, not just individual districts -- although, in practice, a finding of overall unconstitutionality is likely to lead to changes only in the most egregiously gerrymandered districts and some of the districts adjacent to them. Unlike all federal courts that have ruled on partisan gerrymandering to date, the Pennsylvania opinion does not require proof of intent to discriminate, which could make proving a violation easier under state law than under any likely federal standard. Moreover, although the Pennsylvania Supreme Court opinion does not require evidence of partisan effects, and emphasizes failure to comply with good government criteria, we believe that the \textit{LWV} opinion can be ``tweaked" to address what \citet[][see also \citep{Grofman_Cervas_2018_ELJ}]{GrofmanCervas2018_WashPo} in their on-line essay in the \textit{Washington Post} refer to as \textit{stealth gerrymanders}, i.e., plans that satisfy good government redistricting criteria to a reasonable degree but are nonetheless carefully crafted invidious partisan gerrymanders in terms of their expected political impact. The reason for our optimism in this respect are the findings in the opinion that do address partisan effects of the Pennsylvania congressional map. 

@@

    In addition to the rejected 2011 and the court-ordered 2018 map, we also address the partisan and other traditional features of two maps offered to the Pennsylvania Court as a proposed remedy for constitutional violation found in the 2011 map: the \textit{joint legislative} plan offered by Republican legislators from both chambers, and the congressional map proposed by the Democratic governor (\textit{Gov. Wolf}). Here we focus on the question of whether either of these plans was a \textit{stealth gerrymander} done to aid the party drawing the map.

@@

@@

% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
% ▀▄▀▄▀▄ SUBSECTION -- SUBSECTION -- SUBSECTION -- SUBSECTION -- ▄▀▄▀▄▀
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
% ------ ••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
% ↤↤↤↤↤ -------------- S̟U̟B̟-S̟U̟B̟S̟E̟C̟T̟I̟O̟N̟ -------------- ↦↦↦↦↦
% ↤↤↤↤↤ -------------- S̟U̟B̟-S̟U̟B̟S̟E̟C̟T̟I̟O̟N̟ -------------- ↦↦↦↦↦
    Avoiding county splits is incorporated into the language of many state constitutions \citep{Grofman1985} and can be seen as a way of guarding against gerrymandering by making it more difficult to draw partisan maps. There are a total of 67 counties in PA, and the state is allocated 18 congressional districts. Some county splits are necessitated by the size of the county population relative to the ideal population of a congressional district (705,687).  Allegheny and Montgomery must each be split at least twice, while Philadelphia must be split at least three times. If no set of counties exactly equals the ideal congressional population, every districting plan requires a minimum of $n-1$ splits (where $n$ equals the number of congressional districts; assuming that population equality between district must be one person or less). The Pennsylvania Supreme Court found that an unusually and egregiously large number of counties were split in the 2011 congressional map, and they found this fact deeply troubling.

@@

        \footnote{There is a debate about how to count split counties, which was evident when different parties in the case counted different total county splits for the same plan. One way to count splits to count the total number of counties that are in at least two districts. The other way is to count the number of times a county is split, i.e., the number of separate pieces into which the county is divided. In the first counting method, a county can have portions of itself found in 18 different districts, but have this counted as only one county split. Under the second method, this wholesale splitting of the county would be counted as 18 splits. While both methods provide useful information, we regard the second as more informative and it is the one reported in Table 1. However, we must also be alert for divisions of a county into two or more discontiguous pieces found within the same district. We call this “fracking”, to provide a term similar sounding to ‘packing’ and ‘cracking.’ We find “fracking” a particularly egregious problem since there is no need for counties to be fracked, and fracking may be used as a gerrymandering tool. Instances where counties have multiple discontiguous pieces located in a single district are not discussed in the Pennsylvania Supreme Court ruling on the unconstitutionality of the 2011 Pennsylvania congressional plan. However, our analysis of the 2011 plan reveals several cases of fracking: two of the most egregious are districts 14 and 18 in Allegheny County (Pittsburgh). This concept has also been addressed by the U.S. District Court in North Carolina (\textit{Common Cause v. Rucho} No 1:16-CV-1026 (U.S. District Court, Middle District of North Carolina, 2018, slip op at p. 105 [p. 194])).}

@@

% ▀▄▀▄▀▄ F̟I̟G̟U̟R̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ F̟I̟G̟U̟R̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ F̟I̟G̟U̟R̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ F̟I̟
% ---------------------------------------------------------------------
\input{Figures/_maps.tex}
% ---------------------------------------------------------------------
% ▀▄▀▄▀▄ E͎N͎D͎ F͎I͎G͎U͎R͎E͎ ▄▀▄▀▄▀▀▄▀▄▀▄ E͎N͎D͎ F͎I͎G͎U͎R͎E͎ ▄▀▄▀▄▀▀▄▀▄▀▄ E͎N͎D͎ F͎I͎G͎U͎R͎E͎ ▄▀▄

@@

@@

% ↤↤↤↤↤ -------------- S̟U̟B̟-S̟U̟B̟S̟E̟C̟T̟I̟O̟N̟ -------------- ↦↦↦↦↦

@@

    To give a more intuitive feel for how egregiously ill-compact the 2011 map is compared to the other alternatives we are considering, Figure \ref{fig:maps} shows all four maps in the same scale, in a gray and white format. What is visually apparent from these maps is how aesthetically ``ugly" the 2011 map is compared to all of its alternatives. But it also clear that three remedy maps that are similar in having relatively few county cuts nonetheless can look quite different in terms of how each configures districts.

@@

@@

% ▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟ ▄▀▄▀▄▀▀▄▀▄▀▄ T̟A̟B̟L̟E̟
% ---------------------------------------------------------------------
    \input{Tables/_tab_congsum.tex}
        \begin{center}\textbf{INSERT TABLE \ref{tab:congsum} ABOUT HERE} \end{center}

@@

% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
% SUBSECTION
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••
% •••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••••

@@

    There are other indicators of statewide partisan gerrymandering effects that also provide a compelling picture of the 2011 map as an (enduring) partisan gerrymander. There are several metrics that have been proposed to look at the degree to which there is asymmetry in seats-votes relationships that might be indicative of partisan gerrymandering: (1) \textit{partisan bias} \citep{Tufte1973}, (2) the \textit{mean-minus-median} gap \citep{Mcdonald_Best_2015_ELJ} (3) the \textit{efficiency gap} \citep{Stephanopoulos2014_UofChicagoLaw}, and most recently (4) \textit{declination} \citep{Warrington2018}. Each of these measures can be directly calculated for the 2011 map based on district election outcomes. 

@@

        \footnote{Although the basic idea goes back at least as far as \citet{Dahl1956}, the modern \textit{locus classicus} in political science for the idea of partisan bias(and the parallel concept of responsiveness/swing ratio) is \citet{Tufte1973}, with important recent work done by Gary King (see e.g., \citep{GelmanKing1994_unifiedAJPS}; and review in \citet{GrofmanKing2007_ELJ}). A closely related line of research is found in the political geography literature (with \textit{locus classicus} \citet{Brookes1959, Brookes1960} -- with some further methodological improvements by scholars such as \citet{Johnston1994, Rossiter1997} and \citet{Johnston2002}.}   

@@

    \footnote{Where to center the \textit{seats/votes} curve is an unresolved issue in the literature. In \citet{GelmanKing1994_unifiedAJPS} \texttt{JudgeIt} program, the choice is to center the election at the average Democratic vote share over recent elections. \citet{Kastellec_et_al_2008_PS}, similarly, simulates data for average vote shares between 45\% and 55\%. We checked for consistency between the methods by running the numbers in three ways: centering the simulated outcomes at an average election vote share, centering the simulated outcomes at a tied 50/50 election, and centering the simulated outcomes at the actual election results in the particular election. All three variations provide similar estimates, as do the results from \texttt{JudgeIt}. Readers should note that, with only 18 districts, too much weight should not be placed on the exact bias coefficients reported [see the discussion in \citet{Browning_King_1987_seats_votes}]; but the relative magnitude of the bias values across the different plans is fully congruent with the evaluations of probable partisan gerrymandering yielded by the other measures. Because of the limited number of districts in the plans, the same caution is urged for all the measures of gerrymandering reported in this essay.} 

@@

    The \textit{Efficiency Gap} is calculated as defined in \citet{Stephanopoulos2014_UofChicagoLaw}, where all the party’s votes are wasted if they lose the district, and all the winner’s votes over 50\% are wasted. The difference between each party’s wasted votes is then divided by the total votes cast to produce the \textit{efficiency gap}, with a value of zero denoting what is regarded as ideal. As noted in \citet[][p. 13]{Best2018}, this is equivalent to taking an aggregate responsiveness (swing ratio) of two as ideal. 

@@

    Among the more recent measures of gerrymandering is the \textit{Declination}. \textit{Declination} is an astronomical term referring to the angle on a compass, with a Latin root meaning ``a bending away". For gerrymandering, it uses the angles created from ordering all of the districts by vote share, computing the mean vote-share for each party separately for the seats they won, and then comparing the differences in their distribution. To compare, a line is draw from the mean vote-share to the 0.5 line separately for each party. The angle formed from the different slopes of the line is called the \textit{declination}. An angle indicates whether a plan is a gerrymandering, under the logic that vote-shares should be distributed in some uniform manner across districts. If partisan are \textit{packed}, the angle will be greater. All values range between -1 and 1. Though we have defined the vote-share in terms of Republican outcomes, we adjusted the values such that \underline{negative} values are favorable to the Democrats. Further information can be found in \citet{Warrington2018}.  

@@

    Rebuttal to any claims that the partisan asymmetries observed in the consistent outcomes of the 2011 map can be attributed primarily to differential concentration of party voting strengths, or to chance, is found in the expert witness testimony in the case (see esp. the simulations of potential good government maps done by Professor Chen, discussed in detail in the Pennsylvania Supreme Court opinion in the case, and on which the Court placed great reliance). With respect to the durability of the effects of the 2011 map, we have already pointed out that this plan yielded unchanged 13R-5D results for three elections in a row. To help explain this over time consistency we note that the 2011 Map has primarily safe seats (13 of 18 won with greater than a 10\% margin). But, indicative of the pro-Republican gerrymander achieved with the 2011 map, those safe seats are distributed in a highly asymmetric fashion, with nine essentially safe seats for Republicans and only four for Democrats.

@@

@@

%		\footnote{We can only validate our measure for the 2016 election and only for the 2011 enacted plan. For all the remedial plans we are analyzing, the composite is our ``best guess". We have to make some assumptions about state and district inter-election swings. For simplicity, we assume that all districts have a national shift equal to the state-level swing, ie, difference from the 2016 elections. Later in the paper, we conduct simulated elections that vary the district level variability.}

@@

%		\footnote{Whether we take the sum of the five elections or average the five elections, both methods deliver the same relative distributions.}

@@

    Additionally, each district's outcome is identical to the 2016 congressional district results, when we apply a uniform swing to match the results from the state-wide elections to the results from the congressional elections. We are not claiming to predict vote shares in advance of an election. We also recognize that the composite of state-wide elections is intended to provide a ``normal vote" baseline that must be ``corrected" for incumbency effects if we wish to actually predict outcomes.

@@

		\footnote{If we center the simulations at the 2016 statewide Republican vote of 53.3\% of the vote, we again get remarkably consistent estimates [0.09, 0.19, 0.05, 0.46].}

@@

    Such clauses are found in some of the states widely regarded as having the worst congressional plans, such as Maryland and North Carolina.
